---
title: "Supplementary Material"
bibliography: "misc/ref.bib"
csl: "misc/vancouver.csl"
format: 
  docx: 
    reference-doc: "ref.docx"
---

**Title**: Immune-Derived Signatures of RR-Interval Dynamics: Predicting Autonomic Responses from CD21/CD11c B-Cell Subsets

**Authors**: Matías Castillo-Aguilar,^1,2^ ..., Cristian Núñez-Espinosa,^1,2^ \*

::: {custom-style="Afiliation"}
^1^ Centro Asistencial Docente y de Investigación (CADI-UMAG), Punta Arenas, Chile.

^2^ Escuela de Medicina, Universidad de Magallanes (UMAG), Punta Arenas, Chile
:::

### \*Correspondence

Cristian Núñez-Espinosa, Escuela de Medicina, Universidad de Magallanes, Punta Arenas, Chile. Centro Asistencial de Docencia e Investigación CADI-UMAG, Chile. e-mail: [cristian.nunez\@umag.cl](mailto:cristian.nunez@umag.cl). Address: Avenida Bulnes 01855, Box 113-D. Phone: +56 61 2201411.

{{< pagebreak >}}

```{r}
#| include: false

## Load libraries
library(brms)
library(data.table)
library(ggplot2)

## Load auxiliary functions
source("../R/_functions.R")

normalize_text <- function(x) {
    x <- stringi::stri_trans_general(x, "latin-ascii")
    clipr::write_clip(x)
    return(x)
}

## Load data objects
data <- readRDS("../models/m_data.RDS") # Data used in the models
m_sim <- readRDS("../models/m_mod_2.RDS") # Adjusted model
m_adj <- readRDS("../models/m_mod_2.RDS") # Adjusted model
```

In the following sections, we present the parameter effects associated with the neuro-immune interaction and its effects on exercise-induced cardiac autonomic modulation.

## Unadjusted effects

```{r}
data
```


